Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-02-22 parallel-tractability-J-v9/appendix.tex

\textbf{Theorem~\ref{theorem:a1}}
\emph{For any datalog program $P=\langle R, \textbf{I}\rangle$, $P\in\mathcal{D}_{\mathsf{A}_{\text{bsc}}^{\psi}}$ iff
$P$ has a materialization graph whose depth is upper-bounded by $\psi(|P|)$.}\\

\noindent\emph{Proof:} We first prove that the number of iterations of \ref{alg1:updateG}
is actually the depth of the constructed materialization graph.
We define the depths of nodes iteratively as follows:
for each explicit node $v$, \texttt{depth}($v$)=0;
for each implicit node $v'$ whose parents are $v_1,...,v_i$, \texttt{depth}($v'$)=$\max$$\{$ \texttt{depth}($v_1$), ..., \texttt{depth}($v_i$) $\}+1$.
By performing an induction on the number of iterations of \ref{alg1:updateG},
one can check that an implicit node $v'$ has to be added to $\mathcal{G}$ in the $n^{th}$ iteration
where $n$=\texttt{depth}($v'$).
Further, \texttt{depth}($\mathcal{G}$)=$\max$$\{$\texttt{depth}($v_i$)$|v_i$ is in $\mathcal{G}\}$.
We then have that the number of iterations of \ref{alg1:updateG} is \texttt{depth}($\mathcal{G}$).

($\Rightarrow$) $P\in\mathcal{D}_{\mathsf{A}_{\text{bsc}}^{\psi}}$ means that $\mathsf{A}_{\text{bsc}}^{\psi}$ can return
a materialization graph $\mathcal{G}$ of $P$.
Recall that the number of iterations of \ref{alg1:updateG} is bounded by $\psi(|P|)$.
Thus \texttt{depth}($\mathcal{G}$) is also bounded by $\psi(|P|)$.

($\Leftarrow$) Suppose $P$ has a materialization graph $\mathcal{G}$ whose depth
is upper-bounded by $\psi(|P|)$.
If $\mathcal{G}$ has the minimal depth among other materialization graphs of $P$,
the number of iterations of \ref{alg1:updateG} is also \texttt{depth}($\mathcal{G}$) (Lemma~\ref{lemma:a1})
and, thus, upper-bounded by $\psi(|P|)$.
If $\mathsf{A}_{\text{bsc}}$ dose not return $\mathcal{G}$, then the returned graph $\mathcal{G}'$
should have a smaller depth compared with $\mathcal{G}$.
In this case, this conclusion still holds.\hfill$\Box$

@@

\textbf{Lemma~\ref{lemma:a3}}
\emph{Given a datalog program $P=\langle R, \textbf{I}\rangle$,
$\mathsf{A}_{\text{opt}}$ halts and output a materialization graph
$\mathcal{G}$ of $P$.}\\

@@

\emph{Base case}. Initially, $\mathcal{G}$ only contains all the explicit nodes.
In this case, $\mathcal{G}$ is obviously a materialization graph.

@@

\textbf{Lemma~\ref{lemma:dllite}}
\emph{Given a DL-Lite ontology $\mathcal{O}$, for any materialization graph $\mathcal{G}$ of $\mathcal{O}$,
each atom of the form $A(x)$ ($A\neq\bot$) or $R(x,y)$ in
$\mathcal{G}$ has an SWD path.}
\BG{now in the main part:\\ For any DL-Lite$_{\text{core}}$ or DL-Lite$_\mathcal{R}$ ontology $\mathcal{O}$, there exists a materialization graph $\mathcal{G}$ such that
each atom of the form $A(x)$ ($A\neq\bot$) or $R(x,y)$ in
$\mathcal{G}$ has an SWD path.\\ The proof just speaks about DL-Lite
in general as the original claim. Adapt?}\\

\noindent\emph{Proof}: We prove the lemma by an induction on applications of
the datalog rules corresponding to DL-Lite axioms.}\\

\emph{Base case}. For each explicit node $v$ of the form $A(x)$ ($A\neq\bot$) or $R(x,y)$,
$v$ has a special SWD path with $v$ as the unique node. This conclusion holds
for any materialization graph.

\emph{Inductive cases}. For each datalog rule of the form `$B_1,...,B_n\rightarrow H$'
that is rewritten from some DL-Lite axiom, we have the induction hypothesis
that $B_1,...,B_n$ have SWD paths.
We are left to prove that $H$ has an SWD path as well
in all materialization graphs of $\mathcal{O}$.

@@

For both Case~2.1 and Case~2.2, node $R(x,y)$ has an SWD path. Thus,
node $R(x,y)$ has an SWD path in all materialization graphs of $\mathcal{O}$.\hfill$\Box$

@@

\noindent\emph{Proof}: this theorem can be easily proved based on Lemma~\ref{lemma:dllite}.
Specifically, for any DL-lite ontology $\mathcal{O}$,
there always exists a poly-logarithmically bounded function $\psi$ such that
$\mathsf{A}_{\text{opt}}^{\psi}$ can handle the materialization of $\mathcal{O}$,
since in any materialization graph, each node has an SWD path.
More precisely, we can set that $\psi=2$.\hfill$\Box$

@@

\noindent\emph{Proof}: observe that, for a datalog program that is transformed from a DHL ontology,
a rule with a binary atom as its head can only be either of the form (R1), (R2) or (R3).
This also indicates that, the materialization of a DHL
ontology can be separated into two stages: in the first stage (Stage~1), all the rules
of the forms (R1-R3) are exhaustively applied; the consequences of the first stage
also serve as the facts in the second stage (Stage~2). The rules of the forms (T1-T3)
are then applied in Stage~2.
It is obvious that, if both of Stage~1 and Stage~2 can be handled
by performing $\mathcal{A}_{\text{opt}}^\psi$ for some poly-logarithmical function $\psi$, then
the whole materialization can be handled by $\mathcal{A}_{\text{opt}}^\psi$.
In what follows, we investigate the above two stages respectively, and show that,
for any datalog program in $\mathcal{D}_{\textit{\text{dhl}}}$,
such a poly-logarithmical function $\psi$ exists.

In Stage~1, the rules of the forms (R1-R3)
are applied to add new nodes to the constructed materialization graph.
We can observe that, rule (R3) is used for computation of transitive roles.
As mentioned before, there exists an NC algorithm for transitivity computation.
Inspired by this, we can prove that \emph{$\mathcal{A}_{\text{opt}}^\psi$
handles Stage~1 for some poly-logarithmical function $\psi$}.
The proof of this result can be shown by separately considering non-transitive roles and
the roles that are transitive or influenced by transitive roles.
Specifically, we say that role $R$ is \emph{transitively influenced} (TI)
if (1) $R\circ R\sqsubseteq R\in\mathcal{R}$; or
(2) there exists a TI role $R'$ such that $R'\sqsubseteq_*R$ or $R'\sqsubseteq_*R^-$.
We say that a role is an NTI (non-transitively influenced) role if it is not a TI role.
We further define a set $\delta_R$ for each role $R\in\textbf{R}$ as follows:

@@

Note that, for all roles $R$, $\delta_R$ can be computed by only applying (R1) and (R2).
In this sense, Lemma~\ref{lemma:ti} also indicates that:
(1) for each implicit node $R(a,b)$ where $R$ is an NTI role,
it can be added to a materialization graph by only applying (R1) and (R2);
(2) for each transitive role $R$, all implicit nodes of the form $R(a,b)$ are in the transitive closure $\delta^*_{R}$,
which can be computed by an NC algorithm on $\delta_{R}$;
(3) for each role $R$ that is a TI role but not a transitive role,
one can further perform (R1) and (R2) iteratively based on all transitive closures $\delta^*_{R'}$
where $R'$ is a transitive role.
Since all nodes generated by only applying (R1) and (R2) have SWD paths,
the computations of (1) and (3) can be handled by $\mathcal{A}_{\text{opt}}^\psi$.
Further, transitive computation in part (2) can also be handled by $\mathcal{A}_{\text{opt}}^\psi$.
Thus, there exists a poly-logarithmical function $\psi$
such that $\mathcal{A}_{\text{opt}}^\psi$ handles Stage~1.

The results of Stage~1 serve as the facts of Stage~2. In other words, all binary atoms
are explicit nodes in Stage~2. This also means that the rules of the forms (T1) and (T3)
generate nodes with SWD paths. Due to the simple-concept restriction, for each rule
instantiation $A_1(a),A_2(a)\rightarrow B(a)$, one of $A_1(a)$ and $A_2(a)$ always has an SWD path.
Thus, Stage~2 can be handled by $\mathcal{A}_{\text{opt}}$ in at most two iterations.
Based on the above analysis, this theorem holds.

@@

Our target is to show that $\mathcal{A}_{\text{opt}}^\psi$ handles Stage~1. To this end, we also
distinguish all roles by whether they are transitively influenced. Since we have to consider
complex RIAs, we re-define TI and NTI roles as follows:
a role $R$ is \emph{transitively influenced} (TI)
if (1) $R\circ R\sqsubseteq R\in\mathcal{R}$; or
(2) there exists a TI role $R'$ such that $R'\sqsubseteq_*R$ or $R'\sqsubseteq_*R^-$;
or (3) there exist a TI role $R'$ and an axiom of either of the form
$R'\circ R''\sqsubseteq R$ or $R''\circ R'\sqsubseteq R$.
We say that a role is an NTI role if it is not transitively influenced.
The set $\delta_R$ for each role $R\in\textbf{R}$ is re-defined based on \emph{role sequence set}
that is defined as follows:

@@

The set $\delta^*_R$ also denotes the transitive closure of $\delta_R$ where $R$ is a transitive role.
Note that, for all roles $R$, $\delta_R$ can be computed by applying (R1), (R2) and (R4).
The above lemma says:
(1) for each implicit node $R(a,b)$ where $R$ is an NTI role,
it can be added to a materialization graph by only applying (R1), (R2) and (R4);
(2) for transitive roles $R$, all implicit nodes are in the transitive closure $\delta^*_{R}$,
which can be computed by an NC algorithm on $\delta_{R}$;
(3) for each role $R$ that is a TI but not a transitive role,
one can further perform (R1), (R2) and (R4) iteratively based on all
transitive closures $\delta^*_{R'}$ where $R'$ is a transitive role.
All nodes generated by applying (R1) and (R2) have SWD paths in the first iteration; at
least one role in the left hand side of (R4) are restricted to be a simple role.
Thus, the computations of (1) and (3) can be handled by $\mathcal{A}_{\text{opt}}^\psi$.
Further, transitive computation in part (2) can also be handled by $\mathcal{A}_{\text{opt}}^\psi$.
In summary, there exists a poly-logarithmical function $\psi$
such that $\mathcal{A}_{\text{opt}}^\psi$ handles Stage~1.

@@

\textbf{Theorem~\ref{theorem:aprc}}
\emph{For any DHL$(\circ)$ ontology $\mathcal{O}$, Algorithm~$\mathsf{A}_{prc}$ halts and outputs
a materialization graph $\mathcal{O}$.}\\

\noindent\emph{Proof}: similar to the proof of Lemma~\ref{lemma:a3},
this theorem is proved in two stages:
(1) the graph $\mathcal{G}$ returned by $\mathsf{A}_{prc}$ is a materialization graph;
(2) $\mathcal{G}$ is a complete materialization graph.
One can show that (1) holds by performing the same induction on the iterations of \ref{alg3:updateG} of $\mathsf{A}_{\text{opt}}$
(see the proof of Lemma~\ref{lemma:a3}). This is because that
such an induction is performed regardless of how the relation $S_{\textit{\!\tiny rch}}$ is computed.

To prove that (2) holds, we conduct the induction used in the proof for Lemma~\ref{lemma:a1}.
We aim to show that all the ground atoms in $T_R^{i+1}(\textbf{I})$ have to be added to $\mathcal{G}$,
with the induction hypothesis that the ground atoms in $T_R^{i}(\textbf{I})$ are in $\mathcal{G}$.
Suppose the ground atoms in $T_R^{i}(\textbf{I})$ have been added to $\mathcal{G}$ by
performing $\mathsf{A}_{prc}$.
For each atom $\alpha$ in $T_R^{i+1}(\textbf{I})$, there has to be a relation $(\beta,\alpha)$
in $S_{\textit{\!\tiny rch}}$ obtained in some iteration; further
$\beta$ is already in $\mathcal{G}$ according to the inductive hypothesis. Thus,
$\alpha$ has to be added to $\mathcal{G}$ by applying \ref{alg4:updateG}
of $\mathsf{A}_{prc}$.\hfill$\Box$

@@

We first consider Stage~SCM.
Stage~SCM (the simple concept materialization) is conducted on axioms of two forms $A\sqsubseteq B$
and $\exists R.A\sqsubseteq B$, which correspond to datalog rules of the forms $A(x)\rightarrow B(x)$
and $R(x,y),A(y)\rightarrow B(x)$ respectively. Since role assertions are supposed to be fixed
during in Stage~SCM, role $R$ in each rule of the form $R(x,y),A(y)\rightarrow B(x)$
can be viewed as an EDB predicate. This further means that,
for each atom of the form $A(a)$ where $A$ is a simple concept,
$A(a)$ has an SWD path in the first iteration of $\mathsf{A}_{prc}$.
Recall Algorithm~$\mathsf{Prc}$. We use an induction to show that
such an SWD path of $A(a)$ can be determined by $S_{\textit{\!\tiny rch}}$.
The atom $A(a)$ can be derived through either of the following rule instantiations:

@@

We next discuss the non-simple concept materialization (Stage~NSCM), which is conducted on axioms of forms (T1-T3).
According to the simple-concept restriction, for each axiom of the form $A_1\sqcap A_2\sqsubseteq B$,
one of $A_1$ and $A_2$ should be simple concept and can be viewed as an EDB predicate, since Stage~SCM is completed.
In Stage~NSCM, for each atom of the form $A(a)$ where $A$ is not a simple concept,
$A(a)$ has an SWD path in the first iteration of $\mathsf{A}_{prc}$ due to the simple concept restriction.
We also use an induction to show that
such an SWD path of $A(a)$ can be determined by distinguishing different rule instantiations.
The case of the rule instantiation of the form $B(a)\rightarrow A(a)$ (resp., $R(a,b),B(b)\rightarrow A(a)$)
is referred to Case~1 (resp., Case~2). We only consider the case of the rule instantiation of the form $A_1(a),A_2(a)\rightarrow A(a)$
here.

@@

The simple-role materialization (Stage~SRM) can be simply conducted by checking the transitive closure
of role inclusions, i.e., the axioms of the forms (R1-R2).
It is easy to check that the simple-role materialization can be finished after the
first iteration of $\mathsf{A}_{prc}$.

We now discuss Stage~NSRM. For each atom of the form $R(a,b)$ where $R$ is not a simple role
and not a transitive role,
$R(a,b)$ has an SWD path in the first iteration of $\mathsf{A}_{prc}$ due to the simple role restriction.
We use an induction to show that
such an SWD path of $R(a,b)$ can be determined by distinguishing different rule instantiations
as follows.

@@

We finally consider the derivation of the atoms of the form $R(a,b)$, where $R$
is a transitive role.
Recall Lemma~\ref{lemma:tiplus}. We have that, $R(a,b)\in\delta_R$ or
$R(a,b)\in\delta^*_R$ holds.
According to the simple role restriction, for each sub-role $S$ of $R$, $S$ is a simple role,
or there exists such axioms
of the form $S_1\circ S_2\sqsubseteq S$ where $S_1$ and $S_2$ are simple roles.
One can check the set $\delta_R$ can computed by $\mathsf{A}_{prc}$ in
at most two iterations (see Case~4, Case~5 and Case~6).
We should prove that the transitive closure $\delta^*_R$
can also be computed by $\mathsf{A}_{prc}^\psi$ for some
poly-logarithmically bounded function $\psi$.

\begin{enumerate}[leftmargin=12ex]
\item[Case~7] $R(a,c),R(c,b)\rightarrow R(a,b)$ where $R$ is a transitive role.
    One can construct a binary tree $t$ where the root node is $R(a,b)$,
    the leaves are in $\delta_R$ and the hight of $t$ is up-bounded by
    $\log(|\delta_R|)$. It can be checked that, $\mathsf{A}_{prc}$ can generate
    all nodes in each level of $t$ in one iteration from the bottom. Thus,
    $R(a,b)$ can be derived in at most $\log(|\delta_R|)$ iterations.\hfill$\Box$
\end{enumerate}

>>>>

\textbf{Theorem~\ref{theorem:a1}}
\emph{For any datalog program $P=\langle R, \textbf{I}\rangle$, $P\in\mathcal{D}_{\mathsf{A}_{\text{bsc}}^{\psi}}$ iff
$P$ has a materialization graph whose depth is upper-bounded by $\psi(|\textbf{I}|)$.}\\

\noindent\emph{Proof:} We first prove that the number of iterations of \ref{alg1:updateG}
is actually the depth of the constructed materialization graph.
We define the depths of nodes iteratively as follows:
for each of the original node $v$, \texttt{depth}($v$)=0;
for each implicit node $v'$ whose parents are $v_1,...,v_i$, \texttt{depth}($v'$)=$\max$$\{$ \texttt{depth}($v_1$), ..., \texttt{depth}($v_i$) $\}+1$.
By performing an induction on the number of iterations of \ref{alg1:updateG},
one can check that an implicit node $v'$ has to be added to $\mathcal{G}$ in the $n^{th}$ iteration
where $n$=\texttt{depth}($v'$).
Further, \texttt{depth}($\mathcal{G}$)=$\max$$\{$\texttt{depth}($v_i$)$|v_i$ is in $\mathcal{G}\}$.
We then have that the number of iterations of \ref{alg1:updateG} is \texttt{depth}($\mathcal{G}$).

($\Rightarrow$) $P\in\mathcal{D}_{\mathsf{A}_{\text{bsc}}^{\psi}}$ means that $\mathsf{A}_{\text{bsc}}^{\psi}$ can return
a materialization graph $\mathcal{G}$ of $P$.
Recall that the number of iterations of \ref{alg1:updateG} is bounded by $\psi(|\textbf{I}|)$.
Thus \texttt{depth}($\mathcal{G}$) is also bounded by $\psi(|\textbf{I}|)$.

($\Leftarrow$) Suppose $P$ has a materialization graph $\mathcal{G}$ whose depth
is upper-bounded by $\psi(|\textbf{I}|)$.
If $\mathcal{G}$ has the minimal depth among other materialization graphs of $P$,
the number of iterations of \ref{alg1:updateG} is also \texttt{depth}($\mathcal{G}$) (Lemma~\ref{lemma:a1})
and, thus, upper-bounded by $\psi(|\textbf{I}|)$.
If $\mathsf{A}_{\text{bsc}}$ dose not return $\mathcal{G}$, then the returned graph $\mathcal{G}'$
should have a smaller depth compared with $\mathcal{G}$.
In this case, this conclusion still holds.\hfill$\Box$

@@

\textbf{Lemma~\ref{lemma:a3}}
\emph{Given a datalog program $P=\langle R, \textbf{I}\rangle$,
$\mathsf{A}_{\text{opt}}$ halts and outputs a materialization graph
$\mathcal{G}$ of $P$.}\\

@@

\emph{Base case}. Initially, $\mathcal{G}$ only contains all the original nodes.
In this case, $\mathcal{G}$ is obviously a materialization graph.

@@

\textbf{Lemma~\ref{lemma:dllite}}
\emph{For any DL-Lite$_{\text{core}}$ or DL-Lite$_\mathcal{R}$ ontology $\mathcal{O}$, there exists a materialization graph $\mathcal{G}$ such that
each atom of the form $A(x)$ ($A\neq\bot$) or $R(x,y)$ in $\mathcal{G}$ has an SWD path.}\\

\noindent\emph{Proof}: We first prove the lemma by an induction on applications of
the datalog rules corresponding to DL-Lite$_\mathcal{R}$ axioms.\\

\emph{Base case}. For each of the original node $v$ of the form $A(x)$ ($A\neq\bot$) or $R(x,y)$,
$v$ has a special SWD path with $v$ as the unique node. This conclusion holds
for any materialization graph.

\emph{Inductive cases}. For each datalog rule of the form `$B_1,...,B_n\rightarrow H$'
that is rewritten from some DL-Lite$_\mathcal{R}$ axiom, we have the induction hypothesis
that $B_1,...,B_n$ have SWD paths.
We are left to prove that $H$ has an SWD path as well
in all materialization graphs of $\mathcal{O}$.

@@

For both Case~2.1 and Case~2.2, node $R(x,y)$ has an SWD path. Thus,
node $R(x,y)$ has an SWD path in all materialization graphs of $\mathcal{O}$.
We finish the proof for DL-Lite$_\mathcal{R}$. Since 
DL-Lite$_{\text{core}}$ is a subset of DL-Lite$_\mathcal{R}$
in terms of the expressivity, the above conclusion also applies
to DL-Lite$_{\text{core}}$.\hfill$\Box$

@@

\noindent\emph{Proof}: this theorem can be easily proved based on Lemma~\ref{lemma:dllite}.
Specifically, for any DL-Lite$_{\text{core}}$ or DL-Lite$_\mathcal{R}$ ontology $\mathcal{O}$,
there always exists a poly-logarithmically bounded function $\psi$ such that
$\mathsf{A}_{\text{opt}}^{\psi}$ can handle the materialization of $\mathcal{O}$,
since in any materialization graph, each node has an SWD path.
More precisely, we can set that $\psi=2$.\hfill$\Box$

@@

\noindent\emph{Proof}: observe that, for a datalog program that is transformed from a DHL ontology,
a rule with a binary atom as its head can only be either of the form (R1), (R2) or (R3).
This also indicates that, the materialization of a DHL
ontology can be separated into two stages: in the first stage (Stage~1), all the rules
of the forms (R1-R3) are exhaustively applied; the consequences of the first stage
also serve as the facts in the second stage (Stage~2). The rules of the forms (T1-T3)
are then applied in Stage~2.
It is obvious that, if both of Stage~1 and Stage~2 can be handled
by performing $\mathsf{A}_{\text{opt}}^\psi$ for some poly-logarithmical function $\psi$, then
the whole materialization can be handled by $\mathsf{A}_{\text{opt}}^\psi$.
In what follows, we investigate the above two stages respectively, and show that,
for any datalog program in $\mathcal{D}_{\textit{\text{dhl}}}$,
such a poly-logarithmical function $\psi$ exists.

In Stage~1, the rules of the forms (R1-R3)
are applied to add new nodes to the constructed materialization graph.
We can observe that, rule (R3) is used for computation of transitive roles.
As mentioned before, there exists an NC algorithm for transitivity computation.
Inspired by this, we can prove that \emph{$\mathsf{A}_{\text{opt}}^\psi$
handles Stage~1 for some poly-logarithmical function $\psi$}.
The proof of this result can be shown by separately considering non-transitive roles and
the roles that are transitive or influenced by transitive roles.
Specifically, we say that role $R$ is \emph{transitively influenced} (TI)
if (1) $R\circ R\sqsubseteq R\in\mathcal{R}$; or
(2) there exists a TI role $R'$ such that $R'\sqsubseteq_*R$ or $R'\sqsubseteq_*R^-$.
We say that a role is an NTI (non-transitively influenced) role if it is not a TI role.
We further define a set $\delta_R$ for each role $R\in\textbf{R}$ as follows:

@@

Note that, for all roles $R$, $\delta_R$ can be computed by only applying (R1) and (R2).
In this sense, Lemma~\ref{lemma:ti} also indicates that:
(1) for each implicit node $R(a,b)$ where $R$ is an NTI role,
it can be added to a materialization graph by only applying (R1) and (R2);
(2) for each transitive role $R$, all implicit nodes of the form $R(a,b)$ are in the transitive closure $\delta^*_{R}$,
which can be computed by an NC algorithm on $\delta_{R}$;
(3) for each role $R$ that is a TI role but not a transitive role,
one can further perform (R1) and (R2) iteratively based on all transitive closures $\delta^*_{R'}$
where $R'$ is a transitive role.
Since all nodes generated by only applying (R1) and (R2) have SWD paths,
the computations of (1) and (3) can be handled by $\mathsf{A}_{\text{opt}}^\psi$.
Further, transitive computation in part (2) can also be handled by $\mathsf{A}_{\text{opt}}^\psi$.
Thus, there exists a poly-logarithmical function $\psi$
such that $\mathsf{A}_{\text{opt}}^\psi$ handles Stage~1.

The results of Stage~1 serve as the facts of Stage~2. In other words, all binary atoms
are the original nodes in Stage~2. This also means that the rules of the forms (T1) and (T3)
generate nodes with SWD paths. Due to the simple-concept restriction, for each rule
instantiation $A_1(a),A_2(a)\rightarrow B(a)$, one of $A_1(a)$ and $A_2(a)$ always has an SWD path.
Thus, Stage~2 can be handled by $\mathsf{A}_{\text{opt}}$ in at most two iterations.
Based on the above analysis, this theorem holds.

@@

Our target is to show that $\mathsf{A}_{\text{opt}}^\psi$ handles Stage~1. To this end, we also
distinguish all roles by whether they are transitively influenced. Since we have to consider
complex RIAs, we re-define TI and NTI roles as follows:
a role $R$ is \emph{transitively influenced} (TI)
if (1) $R\circ R\sqsubseteq R\in\mathcal{R}$; or
(2) there exists a TI role $R'$ such that $R'\sqsubseteq_*R$ or $R'\sqsubseteq_*R^-$;
or (3) there exist a TI role $R'$ and an axiom of either of the form
$R'\circ R''\sqsubseteq R$ or $R''\circ R'\sqsubseteq R$.
We say that a role is an NTI role if it is not transitively influenced.
The set $\delta_R$ for each role $R\in\textbf{R}$ is re-defined based on \emph{role sequence set}
that is defined as follows:

@@

The set $\delta^*_R$ also denotes the transitive closure of $\delta_R$ where $R$ is a transitive role.
Note that, for all roles $R$, $\delta_R$ can be computed by applying (R1), (R2) and (R4).
The above lemma says:
(1) for each implicit node $R(a,b)$ where $R$ is an NTI role,
it can be added to a materialization graph by only applying (R1), (R2) and (R4);
(2) for transitive roles $R$, all implicit nodes are in the transitive closure $\delta^*_{R}$,
which can be computed by an NC algorithm on $\delta_{R}$;
(3) for each role $R$ that is a TI but not a transitive role,
one can further perform (R1), (R2) and (R4) iteratively based on all
transitive closures $\delta^*_{R'}$ where $R'$ is a transitive role.
All nodes generated by applying (R1) and (R2) have SWD paths in the first iteration; at
least one role in the left hand side of (R4) are restricted to be a simple role.
Thus, the computations of (1) and (3) can be handled by $\mathsf{A}_{\text{opt}}^\psi$.
Further, transitive computation in part (2) can also be handled by $\mathsf{A}_{\text{opt}}^\psi$.
In summary, there exists a poly-logarithmical function $\psi$
such that $\mathsf{A}_{\text{opt}}^\psi$ handles Stage~1.

@@

\textbf{Theorem~\ref{theorem:aprc}}
\emph{For any DHL$(\circ)$ ontology $\mathcal{O}$, Algorithm~$\mathsf{A}_{\text{prc}}$ halts and outputs
a materialization graph $\mathcal{O}$.}\\

\noindent\emph{Proof}: similar to the proof of Lemma~\ref{lemma:a3},
this theorem is proved in two stages:
(1) the graph $\mathcal{G}$ returned by $\mathsf{A}_{\text{prc}}$ is a materialization graph;
(2) $\mathcal{G}$ is a complete materialization graph.
One can show that (1) holds by performing the same induction on the iterations of \ref{alg3:updateG} of $\mathsf{A}_{\text{opt}}$
(see the proof of Lemma~\ref{lemma:a3}). This is because that
such an induction is performed regardless of how the relation $S_{\textit{\!\tiny rch}}$ is computed.

To prove that (2) holds, we conduct the induction used in the proof for Lemma~\ref{lemma:a1}.
We aim to show that all the ground atoms in $T_R^{i+1}(\textbf{I})$ have to be added to $\mathcal{G}$,
with the induction hypothesis that the ground atoms in $T_R^{i}(\textbf{I})$ are in $\mathcal{G}$.
Suppose the ground atoms in $T_R^{i}(\textbf{I})$ have been added to $\mathcal{G}$ by
performing $\mathsf{A}_{\text{prc}}$.
For each atom $\alpha$ in $T_R^{i+1}(\textbf{I})$, there has to be a relation $(\beta,\alpha)$
in $S_{\textit{\!\tiny rch}}$ obtained in some iteration; further
$\beta$ is already in $\mathcal{G}$ according to the inductive hypothesis. Thus,
$\alpha$ has to be added to $\mathcal{G}$ by applying \ref{alg4:updateG}
of $\mathsf{A}_{\text{prc}}$.\hfill$\Box$

@@

We first consider Stage~SCM.
Stage~SCM (the simple concept materialization) is conducted on axioms of two forms $A\sqsubseteq B$
and $\exists R.A\sqsubseteq B$, which correspond to datalog rules of the forms $A(x)\rightarrow B(x)$
and $R(x,y),A(y)\rightarrow B(x)$ respectively. Since role assertions are supposed to be fixed
during in Stage~SCM, role $R$ in each rule of the form $R(x,y),A(y)\rightarrow B(x)$
can be viewed as an EDB predicate. This further means that,
for each atom of the form $A(a)$ where $A$ is a simple concept,
$A(a)$ has an SWD path in the first iteration of $\mathsf{A}_{\text{prc}}$.
Recall Algorithm~$\mathsf{Prc}$. We use an induction to show that
such an SWD path of $A(a)$ can be determined by $S_{\textit{\!\tiny rch}}$.
The atom $A(a)$ can be derived through either of the following rule instantiations:

@@

We next discuss the non-simple concept materialization (Stage~NSCM), which is conducted on axioms of forms (T1-T3).
According to the simple-concept restriction, for each axiom of the form $A_1\sqcap A_2\sqsubseteq B$,
one of $A_1$ and $A_2$ should be simple concept and can be viewed as an EDB predicate, since Stage~SCM is completed.
In Stage~NSCM, for each atom of the form $A(a)$ where $A$ is not a simple concept,
$A(a)$ has an SWD path in the first iteration of $\mathsf{A}_{\text{prc}}$ due to the simple concept restriction.
We also use an induction to show that
such an SWD path of $A(a)$ can be determined by distinguishing different rule instantiations.
The case of the rule instantiation of the form $B(a)\rightarrow A(a)$ (resp., $R(a,b),B(b)\rightarrow A(a)$)
is referred to Case~1 (resp., Case~2). We only consider the case of the rule instantiation of the form $A_1(a),A_2(a)\rightarrow A(a)$
here.

@@

The simple-role materialization (Stage~SRM) can be simply conducted by checking the transitive closure
of role inclusions, i.e., the axioms of the forms (R1-R2).
It is easy to check that the simple-role materialization can be finished after the
first iteration of $\mathsf{A}_{\text{prc}}$.

We now discuss Stage~NSRM. For each atom of the form $R(a,b)$ where $R$ is not a simple role
and not a transitive role,
$R(a,b)$ has an SWD path in the first iteration of $\mathsf{A}_{\text{prc}}$ due to the simple role restriction.
We use an induction to show that
such an SWD path of $R(a,b)$ can be determined by distinguishing different rule instantiations
as follows.

@@

We finally consider the derivation of the atoms of the form $R(a,b)$, where $R$
is a transitive role.
Recall Lemma~\ref{lemma:tiplus}. We have that, $R(a,b)\in\delta_R$ or
$R(a,b)\in\delta^*_R$ holds.
According to the simple role restriction, for each sub-role $S$ of $R$, $S$ is a simple role,
or there exists such axioms
of the form $S_1\circ S_2\sqsubseteq S$ where $S_1$ and $S_2$ are simple roles.
One can check the set $\delta_R$ can computed by $\mathsf{A}_{\text{prc}}$ in
at most two iterations (see Case~4, Case~5 and Case~6).
We should prove that the transitive closure $\delta^*_R$
can also be computed by $\mathsf{A}_{prc}^\psi$ for some
poly-logarithmically bounded function $\psi$.

\begin{enumerate}[leftmargin=12ex]
\item[Case~7] $R(a,c),R(c,b)\rightarrow R(a,b)$ where $R$ is a transitive role.
    One can construct a binary tree $t$ where the root node is $R(a,b)$,
    the leaves are in $\delta_R$ and the hight of $t$ is up-bounded by
    $\log(|\delta_R|)$. It can be checked that, $\mathsf{A}_{\text{prc}}$ can generate
    all nodes in each level of $t$ in one iteration from the bottom. Thus,
    $R(a,b)$ can be derived in at most $\log(|\delta_R|)$ iterations.\hfill$\Box$
\end{enumerate}

### 2018-02-22 parallel-tractability-J-v9/conclusions.tex

In this paper, we studied the problem of parallel tractability of
materialization on the datalog rewritable ontologies.
To identify the parallel tractability classes,
we proposed two \texttt{NC} algorithms, Algorithm~$\mathsf{A}_{bsc}^\psi$
and Algorithm~$\mathsf{A}_{opt}^\psi$, that perform materialization on
datalog rewritable ontology languages.
Based on these algorithms, we identified the corresponding
parallel tractability classes such that materialization
on the datalog programs in these classes is in the NC complexity.
We further studied two specific ontology languages, DL-Lite and DHL (including one of its extension).
We showed that any ontology expressed in DL-Lite$_{core}$ or DL-Lite$_{\mathcal{R}}$ is tractable in parallel.
For DHL and DHL($\circ$), we proposed two restrictions such that materialization is tractable in parallel.

We analyzed the benchmark LUBM and the real-world dataset YAGO
and gave the cases where these ontologies belong to parallel tractability classes.
On the other hand, we used an optimization strategy based on SWD paths
to give a practical algorithm variant Algorithm~$\mathsf{A}_{prc}$, which can
also be restricted to an \texttt{NC} version.
We implemented a system based on Algorithm~$\mathsf{A}_{prc}$ and compared it to
the state-of-the-art reasoner RDFox.
The experimental results showed that our system is comparable to RDFox when handling
the LUBM and the YAGO ontologies.

>>>>

In this paper, we studied the problem of parallel tractability of
materialization on the datalog rewritable ontologies.
To identify the parallel tractability classes,
we proposed two \texttt{NC} algorithms, Algorithm~$\mathsf{A}_{\text{bsc}}^\psi$
and Algorithm~$\mathsf{A}_{\text{opt}}^\psi$, that perform materialization on
datalog rewritable ontology languages.
Based on these algorithms, we identified the corresponding
parallel tractability classes such that materialization
on the datalog programs in these classes is in the NC complexity.
We further studied two specific ontology languages, DL-Lite and DHL (including one of its extension).
We showed that any ontology expressed in DL-Lite$_{core}$ or DL-Lite$_{\mathcal{R}}$ is tractable in parallel.
For DHL and DHL($\circ$), we proposed two restrictions such that materialization is tractable in parallel.

We analyzed the benchmark LUBM and the real-world dataset YAGO
and gave the cases where these ontologies belong to parallel tractability classes.
On the other hand, we used an optimization strategy based on SWD paths
to give a practical algorithm variant Algorithm~$\mathsf{A}_{\text{prc}}$, which can
also be restricted to an \texttt{NC} version.
We implemented a system based on Algorithm~$\mathsf{A}_{\text{prc}}$ and compared it to
the state-of-the-art reasoner RDFox.
The experimental results showed that our system is comparable to RDFox when handling
the LUBM and the YAGO ontologies.
